---
title: "DATA 607 Project 1"
author: "Kristin Lussi"
date: "2023-10-01"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Goal
Your job is to create an R Markdown file that generates a .CSV file (that could for example be imported into a SQL database) with the following information for all of the players:
Player’s Name, Player’s State, Total Number of Points, Player’s Pre-Rating, and Average Pre Chess Rating of Opponents

## Loading and Reading the Data 

First, we will load the packages needed and the url file from github.

```{r loading, warning = FALSE, message = FALSE}
# load packages
library(stringr)
library(dplyr)
library(tidyr)

# specify the url file
urlfile = "https://raw.githubusercontent.com/kristinlussi/607Project1/main/tournamentinfo.txt"

# read the .txt file
chess_data <- read.delim(urlfile, header = FALSE)
```


## Manipulating the Data

Next, we will manipulate the data into a data frame with multiple columns. 

```{r wrangling, warning = FALSE, message = FALSE}
# flatten the data into a vector and then convert into a matrix
chess_data <- matrix(unlist(chess_data), byrow = TRUE)

# remove the "---" rows and create two tables 
# first table has Pair Number, Player Name, Total Points, and Round Scores
chess_data1 <- chess_data[seq(5, length(chess_data), 3)]
# second table has state, USCF #, rating (pre -> post), etc
chess_data2 <- chess_data[seq(6, length(chess_data), 3)]

# change the tables into data frames
chess_data1 <- data.frame(chess_data1)

chess_data2 <- data.frame(chess_data2)

# separate the column into columns: Pair Number (Primary key), Name, 
# Total Points, and each round
chess_data_df <- separate(
  chess_data1,
  col = chess_data1,  
  into = c("Player_Number", "Name", "Total_Points", "Round1", 
           "Round2", "Round3", "Round4", "Round5", "Round6", "Round7"),
  sep = "\\|",
  remove = TRUE 
)

# Change the primary key from character to integer
chess_data_df$Player_Number <- as.integer(chess_data_df$Player_Number)

# separate the column into columns: State, USCF_ID 
# (where the pre_rating is located), and Rating
chess_data_df2 <- separate(
  chess_data2,
  col = chess_data2,  
  into = c("State", "USCF_ID", "RATING"),
  sep = "\\|",  
  remove = TRUE
) 

# create a primary key for the second data frame to match up with the first data frame
chess_data_df2 <- chess_data_df2 %>%
  mutate(Player_Number = row_number()) %>%
  select(Player_Number, everything())

# merge the two data frames by the primary key (Player_Number)
chess_data <- merge(chess_data_df, chess_data_df2, by = "Player_Number")

# separate the USCF_ID column into pre_rating and post_rating columns
chess_data <- chess_data %>%
  separate(USCF_ID, into = c("RATING", "Post_Rating"), sep = " ->", extra = "merge") %>%
  mutate(Post_Rating = gsub(".*R: (\\d+).*", "\\1", Post_Rating)) %>%
  separate(RATING, into = c("RATING", "Pre_Rating"), sep = ": ", extra = "merge") 

# change the Total_Points column to numeric
chess_data$Total_Points <- as.numeric(chess_data$Total_Points)
```


## Calculate Opponent Average Pre-Rating

Here, to assist with calculating the opponent average rating for each player, we will create a second data frame that contains the Player Number and their pre rating to reference back to. We will then make a function which will create a new column for each round, matching each opponent with their corresponding pre rating. These new columns are then used to calculate the mean opponent pre rating for each row (player).

```{r calculate_avg, message = FALSE, warning = FALSE}
# select only the columns needed for opponent average pre rating calculations
chess_data <- chess_data %>% 
  select(Player_Number, Name, State, Total_Points, Pre_Rating, 
         Round1, Round2, Round3, Round4, Round5, Round6, Round7) 

# extract only the player number from each round
chess_data$Round1 <- as.numeric(gsub("[^0-9]+", "", chess_data$Round1))
chess_data$Round2 <- as.numeric(gsub("[^0-9]+", "", chess_data$Round2))
chess_data$Round3 <- as.numeric(gsub("[^0-9]+", "", chess_data$Round3))
chess_data$Round4 <- as.numeric(gsub("[^0-9]+", "", chess_data$Round4))
chess_data$Round5 <- as.numeric(gsub("[^0-9]+", "", chess_data$Round5))
chess_data$Round6 <- as.numeric(gsub("[^0-9]+", "", chess_data$Round6))
chess_data$Round7 <- as.numeric(gsub("[^0-9]+", "", chess_data$Round7))
  
# extract only the first four characters from the Pre_Rating 
# column to get the Pre_Rating for each player
chess_data$Pre_Rating <- substr(chess_data$Pre_Rating, 1, 4) %>%
  as.numeric(chess_data$Pre_Rating) # change the Pre_Rating column to numeric

# create a different data frame for player_numbers and pre_ratings
player_id_rating <- data.frame(chess_data$Player_Number, chess_data$Pre_Rating) 

# rename the colum names in the new data frame
colnames(player_id_rating) <- c("Player_Number", "Opponent_Rating")

# create a function that matches the pre rating to each opponent based on the player number
merge_round_ratings <- function(chess_data, round_column) {
  chess_data %>%
    left_join(player_id_rating, by = setNames("Player_Number", round_column)) %>%
    rename_with(~paste0(round_column, "_Rating"), "Opponent_Rating")
}

# specify each round
rounds <- c("Round1", "Round2", "Round3", "Round4", "Round5", "Round6", "Round7")  

# apply the function for each round
for (round_column in rounds) {
  chess_data <- merge_round_ratings(chess_data, round_column)
}
  
# create a column for average opponent rating and
# calculate the average opponent rating for each row
chess_data <- chess_data %>%
  rowwise() %>%
  mutate(
    Opponent_Avg_Rating = mean(c(Round1_Rating, Round2_Rating, 
                                 Round3_Rating, Round4_Rating, Round5_Rating, 
                                 Round6_Rating, Round7_Rating), na.rm = TRUE)
  )
```    

## Prepare the Final Data Frame for Export

Here, we will select only the Name, State, Total Points, Pre Rating, and Opponent Average Pre Rating columns for our final data frame to prepare for export.

```{r final_data, message = FALSE, warning = FALSE }
# select only the columns required for the .csv file
chess_data <- chess_data %>%
  select(Name, State, Total_Points, Pre_Rating, Opponent_Avg_Rating) %>% 
  as.data.frame()

# show a glimpse of the data frame
head(chess_data)
```

## Exporting the Data into a .csv File

Finally, we can write the .csv file with the following code: 

```{r write_csv, warning = FALSE, message = FALSE}
write.csv(chess_data, file = 
            "/Users/kristinlussi/Documents/DATA607/Project1/chess_data_table.csv", 
          row.names = FALSE, sep = ",")
```

The above .csv file will be saved in the specified file location with the name "chess_data_table.csv". 
